Predicting political party affiliation from text

### 2016-06-13 manuscript/poltext.tex
Changed abstract to match what we actually submitted. removed more text.

\title{Automating Political Bias Prediction}

@@

\begin{abstract}
Every day media generate large amounts of text. An unbiased view on media reports requires an understanding of the political bias of media content. Assistive technology for estimating the political bias of texts can be helpful in this context. This study proposes a simple statistical learning approach to predict political bias from text. Standard text features extracted from speeches and manifestos of political parties are used to predict political bias in terms of political party affiliation and in terms of political views. Results indicate that political bias can be predicted with above chance accuracy. Mistakes of the model can be interpreted with respect to changes of policies of political actors. Two approaches are presented to make the results more interpretable: a) discriminative text features are related to the political orientation of a party and b) sentiment features of texts are correlated with a measure of political power. Political power appears to be strongly correlated with positive sentiment of a text. To highlight some potential use cases a web application shows how the model can be used for texts for which the political bias is not clear such as news articles.
\end{abstract}

@@

\begin{enumerate}
\item {\bf Classification of party affiliation} (five class problem)
\item {\bf Classification of government membership} (binary problem)
\item {\bf Classification of policy issues} (56 class problem)
\end{enumerate}

@@

% Confusion matrix topic level
\begin{table}[t]\label{tab:classification_settings}
\footnotesize
\begin{tabular}{ccc}
Training Set & \multicolumn{2}{c}{Test set} \\
& In-domain&Out-of-domain\\
\hline
\hline
Parliament ($90\%$) & Parliament ($10\%$)& Manifestos\\
Manifestos ($90\%$)& Manifestos ($10\%$) & -\\
\end{tabular}
\end{table}

@@

\section{Classification Model and Training}\label{sec:model}
Bag-of-words feature vectors were used to train a multinomial logistic regression model. Let $y\in\{1,2,\dots,K\}$ be the true  label, where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = &\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}} \qquad \textrm{with }  z_k=&\vec{w}_k^{\top}\vec{x} \\\nonumber
\end{eqnarray}
The performance of the model was optimized using the classification accuracy, but we also report all other standard measures, precision ($TP / (FP + TP$), recall ($TP / (TP + FN)$) and f1-score ($2\times (Prec. \times Rec) / (Prec + Rec.)$). \\

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Wherever possible the generalisation performance of the models was also evaluated on out-of domain texts, for which we used the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).
% Falls wir die Medien nochreinnehmen, muss das bei out-of-domain auch noch erwähnt werden.

\subsection{Sentiment analysis}\label{sec:sentiment_analysis_methods}
A publicly available key word list was used to extract sentiments \cite{remquahey2010}. A sentiment vector $\vec{s}\in\R^d$ was constructed from the sentiment polarity values in the sentiment dictionary. The sentiment index used for attributing positive or negative sentiment to a text was computed as the cosine similarity between BOW vectors $\vec{x}\in\R^d$ and $\vec{s}$

\begin{align}
\frac{\vec{s}^\top \vec{x}}{\|\vec{s}\|\|\vec{x}\|}
\end{align}

@@

\begin{table}[t]
\caption{
\label{tab:results_in-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on in-domain data. $N$ denotes number of data points in the evaluation set.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
       cducsu   &    0.62  &    0.81  &    0.70  &     706\\
        fdp    &   0.70   &   0.37  &    0.49    &   331\\
     gruene &      0.59  &    0.40   &   0.48   &    298\\
      linke    &   0.71   &   0.61  &    0.65    &   338\\
        spd   &    0.60   &   0.69  &    0.65   &    606\\
\hline
avg / total &      0.64   &   0.63   &   0.62    &  2279 
%
\end{tabular}
\end{center}
\end{table}

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
Classification performance on the party affiliation prediction problem for data from the 17th legislative period on out-of-domain data.  Predictions are done on the {\bf sentence level}.\label{tab:confusion}
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 \\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 \\
     gruene   &    0.47    &  0.20   &   0.28    &  3747\\
      linke     &  0.30  &    0.47    &  0.37    &   1701\\
        spd     &  0.26  &    0.16   &   0.20    &   2278\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see  also \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases.}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}

>>>>

\title{Predicting political party affiliation from text}

@@

\begin{abstract}
Every day large amounts of text are produced in public discourse. Some of this text is produced by actors whose political colour is very obvious, for example in the case of party elites. But there are also many actors raising their voices in the discourse, who cannot be clearly attributed to a political party, while their statements might actually by biased towards a specific party. Identifying such biases is crucial for political research as well as media consumers, especially when analysing the influence of the media on political discourse and vice versa. In this study we investigate to what extent political party affiliation can be predicted from textual content. Results indicate that automatic classification of political affiliation is possible with above chance accuracy also across text domains. We propose methods to better interpret these results and find that features not related to political policies, such as speech sentiment, can be discriminative and thus exploited by text analysis models. We present an example of a web application that combines topic modelling with automatic political affiliation analyses of news articles based on the proposed model.
\end{abstract}

@@

\begin{enumerate}
\item {\bf Classification of party affiliation}\\ (five class problem for 17th Bundestag)
\item {\bf Classification of government membership} (binary problem)
\item {\bf Classification of policy issues}\\ (56 class problem)
\end{enumerate}

@@

@@

\section{Classification Model and Training}\label{sec:model}
Bag-of-words feature vectors were used to train a multinomial logistic regression model. Let $y\in\{1,2,\dots,K\}$ be the true  label, where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = \frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}}  \textrm{ with }  z_k=\vec{w}_k^{\top}\vec{x}.
\end{eqnarray}

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of  hyperparameters that were optimised using cross-validation.
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Wherever possible the generalisation performance of the models was also evaluated on out-of domain texts, for which we used the texts from party manifestos. For the policy issue prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation). The train and test splits are summarized in \autoref{tab:classification_settings}.
% Falls wir die Medien nochreinnehmen, muss das bei out-of-domain auch noch erwähnt werden.
% 
\begin{table}[t]\label{tab:classification_settings}
\footnotesize
\begin{tabular}{ccc}
Training Set & \multicolumn{2}{c}{Test set} \\
& In-domain&Out-of-domain\\
\hline
\hline
Parliament ($90\%$) & Parliament ($10\%$)& Manifestos\\
Manifestos ($90\%$)& Manifestos ($10\%$) & -\\
\end{tabular}
\end{table}

\subsection{Sentiment analysis}\label{sec:sentiment_analysis_methods}
A publicly available key word list was used to extract sentiments \cite{remquahey2010}. A sentiment vector $\vec{s}\in\R^d$ was constructed from the sentiment polarity values in the sentiment dictionary. The sentiment index used for attributing positive or negative sentiment to a text was computed as the cosine similarity between BOW vectors and sentiment vectors.

@@

\begin{table}[t]
\caption{
\label{tab:results_in-domain}
{\bf In-domain classification performance} for data from the 17th legislative period on in-domain data. $N$ denotes number of data points in the evaluation set.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
       cducsu   &    0.62  &    0.81  &    0.70  &     706\\
        fdp    &   0.70   &   0.37  &    0.49    &   331\\
     gruene &      0.59  &    0.40   &   0.48   &    298\\
      linke    &   0.71   &   0.61  &    0.65    &   338\\
        spd   &    0.60   &   0.69  &    0.65   &    606\\
\hline
avg / total &      0.64   &   0.63   &   0.62    &  2279 
%
\end{tabular}
\end{center}
\end{table}

\begin{table}[t]
\caption{
\label{tab:results_out-of-domain}
{\bf Out-of-domain classification performance} (quasi-sentence level) of a classifier trained on speeches of the 17th legislative period of the Bundestag.
}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
\hline \hline
    cducsu    &   0.26   &   0.58   &   0.36    &   2030 \\
    fdp    &   0.38   &   0.28   &   0.33    &   2319 \\
     gruene   &    0.47    &  0.20   &   0.28    &  3747\\
      linke     &  0.30  &    0.47    &  0.37    &   1701\\
        spd     &  0.26  &    0.16   &   0.20    &   2278\\
\hline
avg / total    &   0.35  &    0.31  &    0.30   &   12075\\
%
\end{tabular}
\end{center}

@@

\begin{table}[t]
\caption{
\label{tab:results_topic}
{\bf Out-of-domain classification performance} (topic level) on manifesto data of a classifier trained on the 17th legislative period of the Bundestag. Compared to quasi-sentence level predictions (\autoref{tab:results_out-of-domain}) the predictions made on topic level are more reliable.}
\begin{center}
\begin{tabular}{lcccc}
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\end{center}
